Twenty years of dynamic occupancy modelling – what’s been done, and where to next?

### 2024-03-25 Manuscript/domReview.qmd
Restructure
Swapped to a non-standard outline format, away from methods-results-discussion

The dynamic occupancy model (DOM) fills an important niche in the ecological modelling landscape as an approachable midpoint between purely correlative species distributions models and more complex process-based models. Data input requirements are achievable for many research programs, and estimates of occupancy dynamic parameters (i.e., colonisation and extinction) are of use in addressing many key ecological questions. Furthermore, the capacity to control for imperfect detection addresses a ubiquitous issue with field ecological data. A description of the basic model form is presented in **Box 1.**

@@

*INTRO OUTLINE*

drive the past v future angle; how have they been built&used and how can they be in the future?

-   DOMs were developed 20 years ago, they are popular and useful

-   What is a DOM? \[put in figures\]

-   Key assumptions

-   Modelling decisions as they relate to the key assumptions

-   Closure assumption & primary/secondary occasions

-   Heterogeneity and covariates

-   Which covariates should we use?

-   How do we select covariates?

-   & how do we evaluate our models?

-

-   We review articles which fit DOMs

-   Emphasised model building process

-   Synthesis on what further research needs to be done

-   what new potential opportunities exist?

# Methods

# Results

# Discussion

## Use cases

### Observing trends

### Hypothesis testing

### Generating predictions

## Modelling

### Covariates

#### Covariate types

#### Complexity

### Model selection

#### Bayesian implementation

#### Frequentist implementations

### Model evaluation

###

## Future opportunities

### Data types

### Prediction

## Conclusions

>>>>

@@

The dynamic occupancy model (DOM) fills an important niche in the ecological modelling landscape as an approachable midpoint between purely correlative species distributions models and more complex process-based models. Data input requirements are achievable for many research programs, and estimates of occupancy dynamic parameters (i.e., colonisation and extinction) are of use in addressing many key ecological questions. Furthermore, the capacity to control for imperfect detection addresses a ubiquitous source of bias and uncertainty in field ecological data. A description of the basic model form is presented in **Box 1.**

## How are they used

Outputs of dynamic occupancy model may be put to many uses. In many cases, simple estimates of the ecological parameters (i.e. occupancy, colonisation, or extinction) can aid in describing understanding changes in range dynamics. In other cases, estimates of covariates relating to these parameters are used to test hypotheses, for example, the influence of forest fire on the probability of local extinction. Less frequently, DOMs are used to make predictions of site occupancy either across unsurveyed sites or into the future under predictions of future conditions or management actions.

## What does this article do

# History of the dynamic occupancy model

## Precursors

## Variants

## Model extensions

# Review of published implementations

## Systematic review methodology

## Study systems

## Data collection

## Objectives and authorship

## Model development and evaluation

# Future priorities for dynamic occupancy models

## Model selection and evaluation

## Continuous detection data

## Predictions
